Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2014-12-16 ms.tex
updated 3.1

The Nuker profile is a broken power law which transitions from an inner power law slope, $\Gamma$, to an outer power law slope, $\beta$, at a break radius, $\rb$.  Assuming spherical symmetry, this corresponds to a stellar density $\rhostar \sim r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\sim r^{-1-\beta}$ for $r \gg \rb$, thus motivating us to adopt a broken power-law profile
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star(\rsoi) \left(r/\rsoi\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star(r_b) \left(r/r_b\right)^{-1-\beta} & r > r_b,
\end{cases}
\end{align}
where $\rho_\star(\rsoi)$ is the stellar density at the radius of the black hole sphere of influence, 
\be
\rsoi \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6},
\label{eq:rsoi}
\ee
and the final equality in (\ref{eq:rsoi}) employs a recent version of the $\Mbh-\sigma$ relationship (\citealt{Gultekin+09}),
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}

@@

Radiative cooling is not included in the entropy equation, although its effect is assessed critically in $\S\ref{sec:cooling}$.  Across much of the parameter space radiative cooling is negligible compared to other sources of cooling (e.g., advection).  Radiative cooling cannot be neglected in some cases (if
$\vwO$ is low and/or $\eta$ is high), once it becomes important it leads to thermal instabilities, which our one-dimensional time-dependent modeal cannot account for anyways (e.g.~\citealt{McCourt+12}).  A chief goal of this work is instead to use our steady state solutions neglecting radiation to ascertain over what physical conditions cooling instabilities are likely to develop ($\S\ref{sec:cooling}$).

@@

We begin by describing a number of useful analytic results before describing our numerical solutions.

\label{sec:results}
\subsection{Analytics: Flow properties at $\rs$}

This section summarizes analytic estimates of several physical
quantities of interest, such as the stagnation radius $\rs$ and the
mass accretion rate, the detailed derivation of which are given in
Appendix \ref{app:rs}.  All of our results are derived under the
assumption that thermal conductivity can be neglected.

When possible our results are expressed in terms of the wind heating
parameter $\vw$; the power-law slope of the density profile
$n$ (also evaluated at $r = \rs$); Note that $\vw^2=\vwO^2+\sigma^2$,
where the $\sigma$ accounts for shock heating between stellar winds moving
with respect to each other with relative velocity $\sigma$, and $\vwO$
is a free parameter accounting for additional sources of
heating.  $\vw$ may be calculated analytically (for a given $\Mbh$ and
$\vwO$) using $\vw=\chi \vwO$, where

 \begin{align}
&\chi^2 \equiv v_{\rm w}^{2}/v_{\rm w,0}^{2} \approx \left(1 +
  \sigma^{2}/v_{\rm w,0}^{2}\right) \approx \\
&\left(1+0.14 \Mbheight^{0.43}\frac{\vwO}{500 {\rm km/s}}^{-1}\right), 
\end{align}

% where $v_{500} $M_{\bullet,8} \equiv
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$, and the
second equality makes use of the $M_{\bullet}-\sigma$ correlation
(\citealt{Gultekin+09})

 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}

Although the use of equation (\ref{eq:Msigma}) is questionable for low mass BHs (e.g., \citealt{Greene&Ho07}), our results are not particularly sensitive to this assumption.

First, by demanding continuity of the entropy derivative at the stagnation radius, the temperature at the stagnation radius is given by.
\begin{align}
T \simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p v_{w}^{2}}{2} \approx 3.7
\times 10^6\ \, v_{500}^2 \,\,{\rm K} 
\label{eq:Tanalytic}
\end{align}
%{\bf BDM: i picked $\mu = 0.62$ and stick to it.}

Demanding hydrostatic equilibrium at the stagnation point provides an
expression for the stagnation radius

\begin{align}
  \frac{\rs}{\rsoi}=\frac{1}{2 \zeta^2 n}\left[
   \frac{\Mstar(\rs)}{\Mbh}\left(\frac{9-\Gamma}{2}-n\right)+\left(\frac{7}{2}-n\right)\right]
 \label{eq:stag_analytic}
\end{align}

where $\zeta \equiv v_{w,0}/\sigma$, $n$ is the power-law slope of the
density profile at $r = r_{\rm s}$,

\begin{align}
r_{\rm soi} = \frac{GM_{\bullet}}{\sigma^{2}} \approx 14M_{\bullet,8}^{0.6}\,\,{\rm pc}
\end{align}

is the radius of the SMBH sphere of
influence (soi), and A and B are constants that depend on the Nuker
parameter $\Gamma$ and the adiabatic index of the gas.

If the SMBH dominates the potential at the stagnation radius
($M_{\bullet} \gg M_{\star}$; $v_{w,0} \gg 200 M_{\bullet,8}^{0.2}$ km
s$^{-1}$) then the relationship between $\rs$ and $v_{w}$ is
simplified greatly:
\begin{align}
  &\rs=\frac{7}{2}\frac{G \Mbh}{n \vwO^2} \simeq 6
 \, \pc \,\, \Mbheight v_{500}^{-2},
  \label{eq:stag_simple}
\end{align}
where in the numerical estimate we have assume $n = 1$.  This expression is similar to that obtained by \citet{Volonteri+11} on heuristic grounds (their equation~[6]).  On the other hand, if wind heating is much weaker ($v_{\rm w,0} \ll \sigma$), the stagnation radius moves to very large radii or may not exist at all.

The importance of conductive heating can be estimated by comparing its
magnitude to that of wind heating, viz.~
\begin{align} \left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim \frac{2t_{\rm h}\kappa_0
T^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}(r_{\rm s}) v_{\rm w}^{2}}
\nonumber \\ &\sim
53 \eta_{0.02}^{-1}\left(\frac{r_{\rm s}}{r_{\rm
soi}}\right)^{-\beta} v_{500}^{9} \nonumber \\ &\sim
23 \,\, 0.43^{\beta-1} \eta_{0.02}^{-1}
M_{\bullet,8}^{-1.2+0.4\beta} v_{500}^{9-2\beta}
\label{eq:condRatio1}
\end{align}

where we have approximated $\nabla^{2} \sim 1/r_{\rm st}^{2}$ and in
the second equality have made use of equations (\ref{eq:Tanalytic}),
(\ref{eq:stag_simple}) and have approximated $4\pi r_{\rm
  s}^{3}\rho_{\star}(r_{\rm s}) \approx M_{\bullet}(r_{\rm s}/r_{\rm
  soi})^{\beta}$ with $\beta \approx 1-2$.  We thus see that
conductive heating is important near the stagnation radius for high
values of $v_{\rm w} \gtrsim 300$ km s$^{-1}$ and $\eta=0.02$, but
that it is safely neglected to first order for lower heating
rates. {\bf AG:Note that threshold vw is rather--the 300 km/s is
  estimated for cusp galaxies. The change in normalization is due to
  the change in the fiducial eta from 0.1 to 0.02. This seems like bad
  news. Should probably include mass. Although this is tricky as cusp
  and core galaxies have opposite dependencies on mass. Perhaps we
  will have to lean on the saturated argument more if we really would
  like to avoid conductivity. Or perhaps including conductivity is
  truly unavoidable. ):}

In fact, for high heating rates conductive heating is less important
than equation~\eqref{eq:condRatio1} would suggest due to the effects
of saturation. In this regime the importance of conductive heating can
be estimated the conductive flux in equation~\eqref{eq:condRatio1}
with its saturated value.
\begin{eqnarray} \left.\frac{\nabla\cdot(f_{\rm cond})}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim& \frac{2t_{\rm h} \phi \rho(\rs) c_s(\rs)^3}{r_{\rm s}^{2}\eta \rho_{\star}(r_{\rm s}) v_{\rm w}^{2}}
\nonumber \\ &\sim&
\frac{10 \phi}{2-\Gamma}, \label{eq:condRatio}
\end{eqnarray}

where we have approximated $\nabla \sim 1/r_{\rm st}$ and in the
second equality have made use of equations (\ref{eq:Tanalytic}),
(\ref{eq:stag_simple}), and (\ref{eq:rhors}). The stellar density is
assumed to be a power law $\rho_{\star}=\frac{\Mbh
  \left(2-\Gamma\right)}{4 \pi \rsoi^3}
\left(\frac{\rs}{\rsoi}\right)^{-1-\Gamma}$.

From the saturated value of the conductive heat flux we have found an
upper bound on the ratio of the rate of heating to the rate of heat
conduction in equation~\eqref{eq:condRatio}.  For $\phi\simeq 1$, the
conductive flux could be quite important. However, the value of $\phi$
depends various details from plasma physics, and could plausibly be an
order of magnitude smaller. In light of these uncertainties we neglect
the effect of the conductive flux.

The SMBH accretion rate is simply given by the integrated mass loss
rate interior to the stagnation radius (equation~[\ref{eq:q}]), 
%AG: Should we have separate cores and cusps? Or just write a general

\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc(\rs)}{\th} \nonumber \\
&\approx&
  \begin{cases}
    2.9 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    5.3 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}

The Eddington ratio $\eddr$ is given by

\begin{eqnarray}
\eddr&\approx&
  \begin{cases}
    1.3 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core} \\
    2.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}

The stellar density at the stagnation radius (eq.~[\ref{eq:stag_simple}]) is usefully approximated as
\begin{eqnarray}
  && \rho_{\star}(r_{\rm s}) \approx \nonumber \\
  && \begin{cases}
    9.4 \times 10^{-19}M_{\bullet,8}^{-1.2}v_{500}^{2.2}\,{\rm g\,cm^{-3}}
    & \text{core} \\
    1.1 \times 10^{-18} M_{\bullet,8}^{-1.5}v_{500}^{3.6}
    \,{\rm g\,cm^{-3}}  & \text{cusp}, 
  \end{cases}
  \label{eq:rhostarrs}
\end{eqnarray}

The density of gas at the stagnation radius $\rho(r_{\rm s})$ can be roughly estimated by using equations (\ref{eq:stag_simple}) and (\ref{eq:mdot_analytic}) in conjunction with an alternative estimate of the mass accretion rate as the gaseous mass within the stagnation radius divided by the free-fall time $t_{\rm ff}(\rs) = (r_{\rm s}^{3}/GM_{\bullet})^{1/2}$, viz.
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{\tff(\rs)},
  \label{eq:mdot_gas}
\end{align}
from which one derives 
\begin{align}
  \rho(\rs)\simeq
  \begin{cases}
    5 \times 10^{-26} \Mbheight^{-0.2} v_{500}^{-0.8}  \eta_{0.02} \,
    \, {\rm g \, cm^{-3}}& \text{core}\\
    9 \times 10^{-26}  \Mbheight^{-0.5} v_{500}^{0.6}  \eta_{0.02} \,\, {\rm g \,cm^{-3}} & \text{cusp}
  \end{cases}
  \label{eq:rhors}
\end{align}

>>>>

The Nuker profile is a broken power law which transitions from an inner power law slope, $\Gamma$, to an outer power law slope, $\beta$, at a break radius, $\rb$.  Assuming spherical symmetry, this corresponds to a stellar density $\rhostar \sim r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\sim r^{-1-\beta}$ for $r \gg \rb$, thus motivating us to adopt a broken power-law profile
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star(\rsoi) \left(r/\rsoi\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star(r_b) \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star(\rsoi)$ is the stellar density at the radius of the black hole sphere of influence, 
\be
\rsoi \simeq G \Mbh/\sigma^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\ee
and the final equality in (\ref{eq:rsoi}) employs a recent version of the $\Mbh-\sigma$ relationship (\citealt{Gultekin+09}),
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
Although the use of equation (\ref{eq:Msigma}) is questionable for low mass black holes (e.g., \citealt{Greene&Ho07}), our results are not overly sensitive to this assumption.

@@

Radiative cooling is not included in the entropy equation, although its effect is assessed critically in $\S\ref{sec:cooling}$.  Across much of the parameter space radiative cooling is subdominant compared to other sources of cooling (e.g., advection).  Radiative cooling cannot be neglected in some cases (if
$\vwO$ is low and/or $\eta$ is high), once it becomes important it leads to thermal instabilities, which our one-dimensional time-dependent modeal cannot account for anyways (e.g.~\citealt{McCourt+12}).  A chief goal of this work is instead to use our steady state solutions neglecting radiation to ascertain over what physical conditions cooling instabilities are likely to develop ($\S\ref{sec:cooling}$).

@@

We begin by describing useful analytic results before presenting our numerical solutions.

\label{sec:results}
\subsection{Properties near the stagnation radius}

Here we summarize analytic estimates of physical quantities, such as the stagnation radius $\rs$ and the mass accretion rate, the detailed derivation of which are given in Appendix \ref{app:rs}.  Our results are expressed in terms of the SMBH mass $M_{\bullet} = 10^{8}M_{\bullet,8}M_{\odot}$ and the wind heating parameter (eq.~[\ref{eq:dsdt}])
\begin{eqnarray}
v_{\rm w} &=& v_{\rm w,0}\left(1+\frac{\sigma^{2}}{v_{\rm w,0}^{2}}\right)^{1/2} \nonumber \\
 &\approx& v_{\rm w,0}\left(1+0.14 \Mbheight^{0.43}\left[\frac{\vwO}{500\,{\rm km/s}}\right]^{-1}\right)^{1/2},
\end{eqnarray}
where $\vwO$ is a free parameter accounting for sources of heating beyond the baseline level produced by the stellar winds of old stars, and the final equality makes use of $M_{\bullet}-\sigma$ (eq.~[\ref{eq:Msigma}]).

First, continuity of the entropy derivative at the stagnation radius demands that the temperature at the stagnation radius is given by
\begin{align}
T \simeq \frac{\gamma-1}{\gamma}\frac{\mu m_p v_{w}^{2}}{2} \approx 3.7
\times 10^6\ \, v_{500}^2 \,\,{\rm K} 
\label{eq:Tanalytic}
\end{align}
%{\bf BDM: i picked $\mu = 0.62$ and stick to it.}

Hydrostatic equilibrium at the stagnation radius ($v = 0$) can be used to determine the stagnation radius
\begin{align}
  \frac{\rs}{\rsoi}=\frac{\sigma^{2}}{2 v_{w,0}^{2} n}\left[
   \frac{\Mstar(\rs)}{\Mbh}\left(\frac{9-\Gamma}{2}-n\right)+\left(\frac{7}{2}-n\right)\right],
 \label{eq:stag_analytic}
\end{align}
in terms of the radius of the sphere of influence (eq.~[\ref{eq:rsoi}]), where $n$ is the power-law slope of the
density profile at $r = r_{\rm s}$, and A and B are constants that depend on the Nuker
parameter $\Gamma$ and the adiabatic index of the gas.

If the SMBH dominates the gravitational potential at the stagnation radius ($M_{\bullet} \gg M_{\star}$; $v_{w,0} \gg \sigma$) then equation (\ref{eq:stag_analytic}) simplifies greatly
\begin{align}
  &\rs=\frac{7}{2}\frac{G \Mbh}{n \vwO^2} \approx 6
 \, \pc \,\, \Mbheight v_{500}^{-2},
  \label{eq:stag_simple}
\end{align}
where the numerical estimate assumes $n = 1$.  This expression is similar to that obtained by \citet{Volonteri+11} on heuristic grounds (their equation~[6]).  On the other hand, if wind heating is weaker ($v_{\rm w,0} \ll \sigma$), the stagnation radius moves to very large radii or may not exist at all.

Our above analysis neglects heat conduction near the stagnation radius, an assumption we can now check.
The importance of conductive heating can be estimated by comparing its
magnitude to that of wind heating, viz.~
\begin{align}
  \left.\frac{\nabla\cdot(\kappa \nabla T)}{q v_{\rm
w}^{2}/2}\right|_{r_{\rm s}} &\sim \frac{2t_{\rm h}\kappa_0
T^{7/2}}{r_{\rm s}^{2}\eta \rho_{\star}(r_{\rm s}) v_{\rm w}^{2}(1+\phi)}
\nonumber \\ &\sim {\rm min}
  \begin{cases}
  10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.8} v_{500}^{6.8} &  \text{unsaturated (core)}\\
 10 \eta_{0.02}^{-1}
M_{\bullet,8}^{-0.5} v_{500}^{5.4} &  \text{unsaturated (cusp)}\\
  10 \phi/(2-\Gamma) & \text{saturated},
  \end{cases}
 \label{eq:rhors}
\end{align}
where in the second equality have made use of equations (\ref{eq:Tanalytic}), (\ref{eq:stag_simple}) and have approximated $\nabla^{2} \sim 1/r_{\rm st}^{2}$, $\nabla \sim 1/r_{\rm st}$.  Since the stagnation radius is usually well inside the break radius, the stellar density is approximated as (eq.~[\ref{eq:rhostar}])
\begin{eqnarray}
  \rho_{\star}(r_{\rm s}) &\simeq& \frac{M_{\bullet}(2-\Gamma)}{4\pi r_{\rm soi}^{3}}\left(\frac{r_{\rm s}}{r_{\rm soi}}\right)^{-1-\Gamma} \nonumber \\
 &\approx& \begin{cases}
    9.4 \times 10^{-19}M_{\bullet,8}^{-1.2}v_{500}^{2.2}\,{\rm g\,cm^{-3}}
    & \text{core} \\
    1.1 \times 10^{-18} M_{\bullet,8}^{-1.5}v_{500}^{3.6}
    \,{\rm g\,cm^{-3}}  & \text{cusp}, 
  \end{cases}
  \label{eq:rhostarrs}
\end{eqnarray}
Even in the case of saturated heat flux, we see that for expected values $\phi < 1$, heat condution is at most moderately important at the stagnation radius, justifying its neglect in our analysis.

The SMBH accretion rate is simply given by the integrated mass loss
rate interior to the stagnation radius (equation~[\ref{eq:q}]), 
\begin{eqnarray}
  \dot{M} &=& \frac{\eta \Menc(\rs)}{\th} \nonumber \\
&\approx&
  \begin{cases}
    2.9 \times 10^{-5} M_{\bullet,8}^{1.76}
    v_{500}^{-3.8}  \eta_{0.02} \Msun \, \pyear& \text{core} \\
    5.3 \times 10^{-5} M_{\bullet,8}^{1.48} 
    v_{500}^{-2.4}  \eta_{0.02} \Msun \, \pyear  & \text{cusp}, 
  \end{cases}
  \label{eq:mdot_analytic}
\end{eqnarray}
The Eddington ratio $\eddr$ is given by 
\begin{eqnarray}
\eddr&\approx&
  \begin{cases}
    1.3 \times 10^{-5} M_{\bullet,8}^{0.76}
    v_{500}^{-3.8}  \eta_{0.02}   & \text{core} \\
    2.4 \times 10^{-5} \Mbheight^{0.48} 
    v_{500}^{-2.4}  \eta_{0.02}   & \text{cusp}, 
  \end{cases}
  \label{eq:eddr_analytic}
\end{eqnarray}
The density of gas at the stagnation radius $\rho(r_{\rm s})$ is more challenging to estimate.  We roughly estimate its value using equations (\ref{eq:stag_simple}) and (\ref{eq:mdot_analytic}) in conjunction with an alternative estimate of the mass accretion rate as the gaseous mass within the stagnation radius divided by the free-fall time $t_{\rm ff}(\rs) = (r_{\rm s}^{3}/GM_{\bullet})^{1/2}$, viz.
\begin{align}
  &\dot{M}\simeq\frac{(4 \pi/3) \rs^3 \rho(\rs)}{\tff(\rs)},
  \label{eq:mdot_gas}
\end{align}
from which one derives 
\begin{align}
  \rho(\rs)\simeq
  \begin{cases}
    5 \times 10^{-26} \Mbheight^{-0.2} v_{500}^{-0.8}  \eta_{0.02} \,
    \, {\rm g \, cm^{-3}}& \text{core}\\
    9 \times 10^{-26}  \Mbheight^{-0.5} v_{500}^{0.6}  \eta_{0.02} \,\, {\rm g \,cm^{-3}} & \text{cusp}
  \end{cases}
  \label{eq:rhors}
\end{align}
